The Soulgem Paper

### 2011-09-24 soulgem.tex
Second draft of introduction.

People develop software, but the activities that each of them perform to do so differ from those of their colleagues: they divide their labour. For instance, some may write production code, others test it, and still others try to ensure that the product satisfies a real need. In order to talk about this division of labour, we abstract their responsibilities into a number of \emph{roles} (such as ``developer,'' ``tester,'' and ``product manager'') and we use these abstractions as labels to refer to the set of activities that we expect them to carry out. These labels appear everywhere in our academic literature, as well as in books for practitioners, human resources postings, blog posts, and everyday conversation.

Despite this commonplace use, we have no clear indication that the people using these labels use them to refer to the same meanings. For instance, what does a ``developer'' actually need to do in order to fulfill her responsibilities? What should she avoid doing? When should we judge that she has failed to fulfill her responsibilities, and would the answer be generalizable throughout the many domains in which the software industry functions?

In more mature domains, especially in those of critical importance, the division of labour is fairly formal and clear. Surgeons, anesthesiologists, nurses, and patients generally know what to expect from each other, in and out of the operating room. Knowing that someone is ``a nurse'' will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board. In our own community we use role labels just as directly and unthinkingly as in the medical domain, but without realizing (or such is the premise of our paper) that we may be talking about very different things, based on our varying assumptions, experiences, and cultures.

This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has unexplored repercussions for researchers and practitioners alike. Researchers may assume coordination dynamics between, for instance, analysts, designers, and developers that have no connection with the interactions between those groups in reality. Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him. For an abstraction as heavily used as role definitions in software organization, a closer look should prove beneficial.

In this paper we did not attempt to provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success. Instead, (a) we examine the concepts of roles, positions, and expectations in software development groups, providing a theoretical framework based on sociological and organizational research, (b) we report the findings from a case study at two software organizations in which we used this theoretical framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.

>>>>

Poeple divide their labour in order to produce software systems. Some write production code, others test it, and still others try to ensure that the product satisfies a real need. Researchers and practitioners usually abstract these responsibilities into a number of role labels (such as ``developer,'' ``tester,'' and ``product manager''); we use these abstractions to refer to the set of activities that we expect them to carry out. These labels are common everywhere in our academic literature, as well as in textbooks, books for practitioners, human resources postings, blog posts, and everyday conversation.

Despite this commonplace use, we have no clear indication that the people using these labels refer to the same meanings. For instance, what does a ``developer'' actually need to do in order to fulfill her responsibilities? What should we expect her to be unable to do? And would the answer be generalizable throughout the many areas in which the software industry operates?

In some domains, especially the more mature or critical ones, the division of labour is fairly formal and clear. Surgeons, anesthesiologists, nurses, and patients generally know what to expect from each other, in and out of the operating room. Knowing that someone is a nurse will bring out a set of expectations of behaviour regarding that person that have a good probability of being shared across the board. But although in our own community we assign labels just as directly and unthinkingly as in the medical domain, we do not realize (or such is the premise of our paper) that we may be talking about very different things, based on our varying assumptions, experiences, and cultures.

This ontological drift \cite{Robinson1991} in our conceptualization of software development roles has unexplored repercussions for researchers and practitioners alike. Researchers may prescribe practices to be carried out by one role without realizing that the role is usually shared and spread out among a group of people. Practitioners may not know how a newcomer's experience as a developer in a different organization will map into their own, or how to train him. For an abstraction as heavily used as role definitions in software organization, a closer look should prove beneficial.

In this paper we do not attempt to provide a rigid definition of roles in software organizations; in fact we argue that such a task has poor prospects of success. Instead, (a) we examine the concepts of roles, positions, and expectations in software development groups, providing a theoretical framework based on sociological and organizational research, (b) we report the findings from a case study at two software organizations in which we used this theoretical framework to ground our observations and interviews, and (c) we discuss the implications for research and practice from the framework and from our findings.
